---
title: "Predicting GDP growth for selected countries"
output:
  html_document: default
  pdf_document: default
date: '2024-01-14'
---

## Data Structure

A dataset containing 1311 entries on 6 variables, for 57 countries over 23 years. The variables are as follows:

-   **Year**: The year of observation
-   **GDP Growth**: Annual GDP growth rate
-   **GFCF change**: Gross Fixed Capital Formation change
-   **Inflation change**: Percentage change in inflation
-   **Unemployment change**: Percentage change in unemployment (some entries may have null values)

## Objective

Predict the GDP growth rate with the other variables as predictors.

------------------------------------------------------------------------

**Loading the data and necessary libraries**

```{r}
# Load necessary libraries
library(tidyverse)
library(viridis)
library(car)
library(lubridate)
library(broom)
library(ggplot2)
library(stats)
library(stargazer)
library(dplyr)
library(lmtest)
library(sandwich)
library(forecast)
library(stats)
library(reshape2)
library(gridExtra)
library(plm)
library(broom)
library(gt)

# Read the data
data <- read.csv('economic_indicators.csv')
```

## Contents

**1. Data Exploration**

**2. Pooled OLS estimation**

**3. Fixed effects estimation**

**4. Random effects estimation**

------------------------------------------------------------------------

### 1. Data Exploration

-   Data Stats
-   Checking variable correlations
-   Checking the Variance Inflation Factor (VIF)
-   Looking at the data

```{r}
head(data)
```

```{r}
num_unique_years <- length(unique(data$Year))
print(paste("Number of unique years: ", num_unique_years))

num_unique_countries <- length(unique(data$Country.Name))
print(paste("Number of unique countries: ", num_unique_countries))

num_data_points <- length(data$Country.Name)
print(paste("Total number of data: ", num_data_points))
```

```{r}
# Ensure only numeric columns are included for correlation matrix
numeric_data_explore <- data %>% select_if(is.numeric) %>%
  select(-Year)

# Compute the correlation matrix
correlation_matrix <- cor(numeric_data_explore)

# Convert the correlation matrix to a long format
long_corr <- melt(correlation_matrix)

# Create a heatmap
ggplot(long_corr, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", value)),
            size = 3, color = "white", vjust = 1) + 
  scale_fill_viridis(option='viridis') +
  theme_minimal() +
  coord_fixed() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap", x = NULL, y = NULL)

```

```{r}
model_vif <- lm(GDP.Growth ~ Inflation.change + Unemployment.change + GFCF.change, data = data)
vif_vals <- vif(model_vif)
barplot(vif_vals, main="VIF Values", horiz= TRUE, col = 'steelblue')
```

**Scatter plots**

```{r}
# Extract the correlations with 'GDP.Growth'
gdp_corrs <- correlation_matrix["GDP.Growth",]

# Create scatterplots
plot1 <- ggplot(numeric_data_explore, aes(x = `GDP.Growth`, y = `GFCF.change`)) +
  geom_point() +
  labs(title = paste('Correlation:', round(gdp_corrs["GFCF.change"], 2)))

plot2 <- ggplot(numeric_data_explore, aes(x = `GDP.Growth`, y = `Inflation.change`)) +
  geom_point() +
  labs(title = paste("Correlation:", round(gdp_corrs["Inflation.change"], 2)))

plot3 <- ggplot(numeric_data_explore, aes(x = `GDP.Growth`, y = `Unemployment.change`)) +
  geom_point() +
  labs(title = paste("Correlation:", round(gdp_corrs["Unemployment.change"], 2)))

# Arrange the plots in a grid
grid.arrange(plot1, plot2, plot3, nrow = 1)

```

Capping the inflation change at a maximum of 5%, by removing countries with such inflations.

```{r}
# Cap the inflation at -5, 5 and exclude countries with Inflation change greater than 5
data_capped <- ungroup(data %>%
 group_by(Country.Name) %>%
 filter(all(abs(Inflation.change) <= 5)))

data_capped_n <- select(data_capped, GDP.Growth, GFCF.change, Inflation.change, Unemployment.change)


# Compute the correlation matrix for the capped data
corrs <- cor(data_capped_n)

# Create scatterplots
plot1 <- ggplot(data_capped_n, aes(x = `GDP.Growth`, y = `GFCF.change`)) +
  geom_point() +
  labs(title = paste("Correlation:", round(corrs["GDP.Growth", "GFCF.change"], 2)))

plot2 <- ggplot(data_capped_n, aes(x = `GDP.Growth`, y = `Inflation.change`)) +
  geom_point() +
  labs(title = paste("Correlation:", round(corrs["GDP.Growth", "Inflation.change"], 2)))

plot3 <- ggplot(data_capped_n, aes(x = `GDP.Growth`, y = `Unemployment.change`)) +
  geom_point() +
  labs(title = paste("Correlation:", round(corrs["GDP.Growth", "Unemployment.change"], 2)))

# Arrange the plots in a grid
grid.arrange(plot1, plot2, plot3, nrow = 1)

```

Lets now check the stats about our data

```{r}
data <- data_capped

num_unique_years <- length(unique(data$Year))
print(paste("Number of unique years: ", num_unique_years))

num_unique_countries <- length(unique(data$Country.Name))
print(paste("Number of unique countries: ", num_unique_countries))

num_data_points <- length(data$Country.Name)
print(paste("Total number of data: ", num_data_points))
```

### 2. Pooled OLS

------------------------------------------------------------------------

#### Fitting the model

```{r}

X <- data %>% 
  select(`GFCF.change`, `Unemployment.change`, `Inflation.change`)


y <- data$`GDP.Growth`


# Fit OLS model
pooled_ols <- lm(y ~ ., data = X)

# Print summary table
summary(pooled_ols)
```

#### Are residuals normally distributed?

-   Looking at the Jarque-Bera and omnibus tests, we have non normality in the residuals
-   This doesn't mean that the estimator isn't the BLUE (!)

```{r}

# Create Q-Q plot of the residuals with the reference line
qq_plot <- ggplot() + 
  stat_qq(aes(sample = residuals(pooled_ols))) + 
  stat_qq_line(aes(sample = residuals(pooled_ols)), color = "red") +
  ggtitle("Normal Q-Q") +
  theme_minimal()

# Create Kernel Density Plot of the residuals
kde_plot <- ggplot(data = data.frame(resid = residuals(pooled_ols)), aes(x = resid)) +
  geom_density() +
  stat_function(fun = dnorm, args = list(mean = mean(residuals(pooled_ols)), 
                                         sd = sd(residuals(pooled_ols))), 
                colour = "red", linetype = "dashed") +
  ggtitle("Residuals KDE") +
  labs(x = "Residuals", y = "Density") +
  theme_minimal()

# Arrange the plots in a grid
grid.arrange(qq_plot, kde_plot, nrow = 1)
```

#### Are the residuals homoskedastic?

```{r}
# Plot residuals vs fitted values
ggplot() +
  geom_point(aes(x = fitted(pooled_ols), y = residuals(pooled_ols)), color = 'blue') +
  labs(x = 'Fitted Values', y = 'Residuals', title = 'Residuals vs Fitted Values') +
  theme_minimal()
```

From the plot we couldn't be sure. Lets use the breusch pagan and white test to check

```{r}
# Run White's test for heteroskedasticity
white_test <- bptest(pooled_ols, ~ fitted(pooled_ols) + I(fitted(pooled_ols)^2), data = data)

# Run Breusch-Pagan test for heteroskedasticity
bp_test <- bptest(pooled_ols)

# Output the p-values
cat('White Heteroskedasticity Test p-value:', white_test$p.value, '\n')
cat('Breusch Pagan Heteroskedasticity Test p-value:', bp_test$p.value, '\n')

```

Clearly there is heteroskedasticity -\> Gauss-Markov assumptions are violated -\> OLS isnt the BLUE

#### Are the residuals correlated with the endogenous variable? Lets visually inspect (exogeneity)

```{r warning=False, message=False}
# Scatterplot of residuals vs GDP Growth
correlation <- cor(data$`GDP.Growth`, residuals(pooled_ols))
ggplot(data = data, aes(x = `GDP.Growth`, y = residuals(pooled_ols))) +
 geom_point() +
 geom_smooth(method = "lm", se = FALSE, color = "red") +
 labs(x = 'GDP Growth', y = 'Residuals', title = paste("Correlation: ", round(correlation, 2))) +
 theme_minimal()
```

Clearly they are very correlated.

#### Are the residuals serially autocorrelated?

```{r}
# Load necessary libraries

# Create a time series object for residuals
residuals_ts <- ts(residuals(pooled_ols))

# Create subplots for ACF and PACF
par(mfrow=c(1,2))
acf(residuals_ts, main="Autocorrelation", lag.max=20)
pacf(residuals_ts, main="Partial Autocorrelation", lag.max=20)
```

Definetely autocorrelation. Lets move on to the fixed effects estimator.

### 3. Fixed Effects

------------------------------------------------------------------------

```{r warning=False, message=False}
library(plm)
fe <- plm(GDP.Growth ~ GFCF.change + Inflation.change + Unemployment.change, model='within', data=data)
summary(fe)
```

#### Is the fixed effects model better than the pooled ols? Let's run an f-test

```{r}
pFtest(fe, pooled_ols)
```

Definetely better than the pooled ols !

### 4. Random Effects

------------------------------------------------------------------------

```{r}
re <- plm(GDP.Growth ~ GFCF.change + Inflation.change + Unemployment.change, model="random", data=data)
summary(re)
```
#### Should we use fixed or random effects ? (Hausman test)

```{r}
phtest(fe, re)
```
We should definetely use fixed effects instead, as the c_i terms (individual specific intercepts) and regressors are correlated. There is also heterosckedasticity, as seen with the pooled OLS, so the 


